Aligning marine species range data to better serve science and conservation

### 2016-08-17 case_study_ohi/ohi_single_datasets.Rmd

### Load species cells; filter to species IDs in assessed species list and cell IDs in OHI regions
am_spp_cells <- read_csv(file.path(dir_anx, 'v2016/int/am_cells_spp_prob0.csv')) %>%
  filter(am_sid %in% ohi_spp_info$am_sid) %>% 
  filter(loiczid %in% ohi_rgn_cells$loiczid)

### bind IUCN loiczid dataframe to species risk category dataframe
am_spp_cells <- am_spp_cells %>%
  left_join(ohi_spp_info %>% select(am_sid, cat_score),
            by = 'am_sid')

iucn_spp_cells <- read_csv(file.path(dir_anx, 'v2016/int/iucn_cells_spp.csv'),
                           col_types = 'cddddc') %>%
  filter(presence != 5) %>%
  filter(iucn_sid %in% ohi_spp_info$iucn_sid) %>%
  filter(loiczid %in% ohi_rgn_cells$loiczid)
### bind IUCN loiczid dataframe to species risk category dataframe
iucn_spp_cells <- iucn_spp_cells %>%
  left_join(ohi_spp_info %>% select(iucn_sid, cat_score),
            by = 'iucn_sid')

@@

### group by loiczid, then summarize to find means (and counts)
iucn_cell_sum <- iucn_spp_cells %>%
  filter(!is.na(cat_score)) %>%
  group_by(loiczid) %>%
  summarize(mean_cat = mean(cat_score),
            n_spp    = n())

@@

### bind cells to regions dataframe and summarize
iucn_rgn_sum <- iucn_cell_sum %>%
  inner_join(ohi_rgn_cells,
             by = 'loiczid') %>%
  # mutate(rgn_area = cell_area * proportionArea,
  #        area_weighted_mean_cat   = weighted_mean_cat   * rgn_area,
  #        area_weighted_mean_trend = weighted_mean_trend * rgn_area) %>%
  mutate(cell_area_weight = cell_area * n_spp * proportionArea,
         area_weighted_mean_cat = mean_cat * cell_area_weight) %>%
  arrange(loiczid)

iucn_rgn_sum <- iucn_rgn_sum %>%
  group_by(rgn_id) %>%
  summarize(rgn_mean_cat = sum(area_weighted_mean_cat)/sum(cell_area_weight),
            mean_n_spp = mean(n_spp))

@@

### group by loiczid, then summarize to find means (and counts)
am_cell_sum <- am_spp_cells %>%
  filter(!is.na(cat_score)) %>%
  group_by(loiczid) %>%
  summarize(mean_cat = mean(cat_score),
            n_spp    = n())

@@

### bind cells to regions dataframe and summarize
am_rgn_sum <- am_cell_sum %>%
  inner_join(ohi_rgn_cells,
             by = 'loiczid') %>%
  # mutate(rgn_area = cell_area * proportionArea,
  #        area_weighted_mean_cat   = weighted_mean_cat   * rgn_area,
  #        area_weighted_mean_trend = weighted_mean_trend * rgn_area) %>%
  mutate(cell_area_weight = cell_area * n_spp * proportionArea,
         area_weighted_mean_cat = mean_cat   * cell_area_weight) %>%
  arrange(loiczid)

am_rgn_sum <- am_rgn_sum %>%
  group_by(rgn_id) %>%
  summarize(rgn_mean_cat = sum(area_weighted_mean_cat)/sum(cell_area_weight),
            mean_n_spp = mean(n_spp))

@@

am_cell_sum_both <- am_spp_cells %>%
  filter(am_sid %in% spp_ohi_am$am_sid) %>%
  filter(!is.na(cat_score)) %>%
  group_by(loiczid) %>%
  summarize(mean_cat = mean(cat_score),
            n_spp    = n(),
            data     = 'am')

@@

iucn_cell_sum_both <- iucn_spp_cells %>%
  filter(iucn_sid %in% spp_ohi_iucn$iucn_sid) %>%
  filter(!is.na(cat_score)) %>%
  group_by(loiczid) %>%
  summarize(mean_cat = mean(cat_score),
            n_spp    = n(),
            data     = 'iucn')

@@

ohi_rgn_sum <- ohi_rgn_sum %>%
  group_by(rgn_id) %>%
  summarize(rgn_mean_cat = sum(area_weighted_mean_cat)/sum(cell_area_weight),
            mean_n_spp = mean(n_spp))

@@

status_df <- ohi_rgn_sum %>%
  select(rgn_id, ohi_st = status, mean_n_spp) %>%
  mutate(ohi_st = ohi_st) %>%
  full_join(am_rgn_sum   %>% select(rgn_id,    am_st = status), by = 'rgn_id') %>%
  full_join(iucn_rgn_sum %>% select(rgn_id, iucn_st = status), by = 'rgn_id') %>%
  mutate(diff_am   = am_st - ohi_st ,
         diff_iucn = iucn_st - ohi_st) %>%
  gather(scenario, diff, diff_am:diff_iucn)

@@

>>>>

### Load species cells; filter to species IDs in assessed species list and cell IDs in OHI regions
am_spp_cells <- read_csv(file.path(dir_anx, 'v2016/int/am_cells_spp_prob0.csv')) %>%
  filter(am_sid %in% ohi_spp_info$am_sid) %>% 
  filter(loiczid %in% ohi_rgn_cells$loiczid) %>%
  distinct()

### bind AM loiczid dataframe to species risk category dataframe
am_spp_cells <- am_spp_cells %>%
  left_join(ohi_spp_info %>% select(am_sid, cat_score),
            by = 'am_sid')

### force exclusion of bird IUCN SIDs for IUCN analysis only
iucn_nonbirds <- ohi_spp_info %>%
  select(iucn_sid, spp_group) %>%
  filter(spp_group != 'BOTW')

iucn_spp_cells <- read_csv(file.path(dir_anx, 'v2016/int/iucn_cells_spp.csv'),
                           col_types = 'cddddc') %>%
  filter(presence != 5) %>%
  filter(iucn_sid %in% ohi_spp_info$iucn_sid) %>%
  filter(iucn_sid %in% iucn_nonbirds$iucn_sid) %>% ### exclude BOTW
  filter(loiczid %in% ohi_rgn_cells$loiczid) %>%
  distinct()
### bind IUCN loiczid dataframe to species risk category dataframe
iucn_spp_cells <- iucn_spp_cells %>%
  left_join(ohi_spp_info %>% select(iucn_sid, cat_score),
            by = 'iucn_sid')

@@

### group by loiczid, then summarize to find means (and counts)
iucn_cell_sum <- iucn_spp_cells %>%
  filter(!is.na(cat_score)) %>%
  distinct() %>%
  group_by(loiczid) %>%
  summarize(mean_cat = mean(cat_score),
            n_spp    = n())

@@

### bind cells to regions dataframe and summarize
iucn_rgn_sum <- iucn_cell_sum %>%
  inner_join(ohi_rgn_cells,
             by = 'loiczid') %>%
  mutate(cell_area_weight = cell_area * n_spp * proportionArea,
         area_weighted_mean_cat = mean_cat * cell_area_weight) %>%
  arrange(loiczid)

iucn_rgn_sum <- iucn_rgn_sum %>%
  group_by(rgn_id, rgn_name) %>%
  summarize(rgn_mean_cat = sum(area_weighted_mean_cat)/sum(cell_area_weight),
            mean_n_spp = mean(n_spp))

@@

### group by loiczid, then summarize to find means (and counts)
am_cell_sum <- am_spp_cells %>%
  filter(!is.na(cat_score)) %>%
  distinct() %>%
  group_by(loiczid) %>%
  summarize(mean_cat = mean(cat_score),
            n_spp    = n())

@@

### bind cells to regions dataframe and summarize
am_rgn_sum <- am_cell_sum %>%
  inner_join(ohi_rgn_cells,
             by = 'loiczid') %>%
  mutate(cell_area_weight = cell_area * n_spp * proportionArea,
         area_weighted_mean_cat = mean_cat   * cell_area_weight) %>%
  arrange(loiczid)

am_rgn_sum <- am_rgn_sum %>%
  group_by(rgn_id, rgn_name) %>%
  summarize(rgn_mean_cat = sum(area_weighted_mean_cat)/sum(cell_area_weight),
            mean_n_spp = mean(n_spp))

@@

am_cell_sum_both <- am_spp_cells %>%
  filter(am_sid %in% spp_ohi_am$am_sid) %>%
  filter(!is.na(cat_score)) %>%
  distinct() %>%
  group_by(loiczid) %>%
  summarize(mean_cat = mean(cat_score),
            n_spp    = n(),
            data     = 'am')

@@

iucn_cell_sum_both <- iucn_spp_cells %>%
  filter(iucn_sid %in% spp_ohi_iucn$iucn_sid) %>%
  filter(!is.na(cat_score)) %>%
  distinct() %>%
  group_by(loiczid) %>%
  summarize(mean_cat = mean(cat_score),
            n_spp    = n(),
            data     = 'iucn')

@@

ohi_rgn_sum <- ohi_rgn_sum %>%
  group_by(rgn_id, rgn_name) %>%
  summarize(rgn_mean_cat = sum(area_weighted_mean_cat)/sum(cell_area_weight),
            mean_n_spp = mean(n_spp))

@@

status_df <- ohi_rgn_sum %>%
  select(rgn_id, rgn_name, ohi_st = status, mean_n_spp) %>%
  distinct() %>%
  mutate(ohi_st = ohi_st) %>%
  full_join(am_rgn_sum   %>% 
              select(rgn_id, am_st = status) %>%
              distinct(), 
            by = 'rgn_id') %>%
  full_join(iucn_rgn_sum %>% 
              select(rgn_id, iucn_st = status) %>%
              distinct(), 
            by = 'rgn_id') %>%
  mutate(diff_am   = am_st - ohi_st ,
         diff_iucn = iucn_st - ohi_st) %>%
  gather(scenario, diff, diff_am:diff_iucn)

@@

``` {r check_weirdness}
### noting that the mean and median by AM only and by IUCN only are both less than OHI...

status_df$am_st %>% mean()     ### 91.352
status_df$am_st %>% median()   ### 92.164
status_df$iucn_st %>% mean()   ### 90.943
status_df$iucn_st %>% median() ### 92.565
status_df$ohi_st %>% mean()    ### 93.628
status_df$ohi_st %>% median()  ### 93.803

am_cell_sum_all <- read_csv(file.path(dir_git, 'case_study_ohi/am_spp_cell_sum.csv'))
iucn_cell_sum_all <- read_csv(file.path(dir_git, 'case_study_ohi/iucn_spp_cell_sum.csv'))
ohi_cell_sum_test <- iucn_cell_sum_both %>%
  rename(mean_cat_i = mean_cat, n_spp_i = n_spp) %>%
  select(-data) %>%
  full_join(am_cell_sum_both %>%
              rename(mean_cat_a = mean_cat, n_spp_a = n_spp) %>%
              select(-data), 
            by = 'loiczid')

spp_info_trunc <- ohi_spp_info %>%
  select(am_sid, iucn_sid, sciname, cat_score, spp_group, spatial_source)

spp_ohi_am <- spp_info_trunc %>% 
  filter(str_detect(spatial_source, 'am'))
spp_ohi_iucn <- spp_info_trunc %>% 
  filter(str_detect(spatial_source, 'iucn'))

### examine one cell with lots of species to see what's happening
a_all <- am_spp_cells %>%
  filter(loiczid == 161563) %>%
  inner_join(spp_info_trunc) %>%
  distinct()
a_ohi <- am_spp_cells %>%
  filter(loiczid == 161563) %>%
  inner_join(spp_ohi_am) %>%
  mutate(type = 'ohi') %>%
  distinct()
a <- a_all %>%
  full_join(a_ohi)
i_all <- iucn_spp_cells %>%
  filter(loiczid == 161563) %>%
  inner_join(spp_info_trunc) %>%
  distinct()
i_ohi <- iucn_spp_cells %>%
  filter(loiczid == 161563) %>%
  inner_join(spp_ohi_iucn) %>%
  mutate(type = 'ohi') %>%
  distinct()
i <- i_all %>%
  full_join(i_ohi)

# > mean(a$cat_score)
# [1] 0.04671916
# > mean(a$cat_score[is.na(a$type)])
# [1] 0.2727273
### for OHI, 44 (out of 381) relatively unhealthy species are removed from AM presence for this cell,
### because the presence is driven by IUCN maps.
### If they aren't replaced by IUCN presence, then this cell's score will seem artificially
### healthier.

### OK check a region with large IUCN changes and small spp numbers:
### rgn 4 Macquarie Island

macq_cells <- ohi_rgn_cells %>% filter(rgn_id == 4)
### examine one rgn with lots of species to see what's happening
a_rgn_all <- am_spp_cells %>%
  filter(loiczid %in% macq_cells$loiczid) %>%
  inner_join(spp_info_trunc) %>%
  distinct() %>%
  group_by(am_sid, cat_score, sciname) %>%
  summarize(n_cells_spp = n())
a_rgn_ohi <- am_spp_cells %>%
  filter(loiczid %in% macq_cells$loiczid) %>%
  inner_join(spp_ohi_am) %>%
  distinct() %>%
  group_by(am_sid, cat_score, sciname) %>%
  summarize(n_cells_spp = n()) %>%
  mutate(type = 'ohi')
a_rgn <- a_rgn_all %>%
  full_join(a_rgn_ohi)
i_rgn_all <- iucn_spp_cells %>%
  filter(loiczid %in% macq_cells$loiczid) %>%
  inner_join(spp_info_trunc) %>%
  distinct() %>%
  group_by(iucn_sid, cat_score, sciname) %>%
  summarize(n_cells_spp = n())
i_rgn_ohi <- iucn_spp_cells %>%
  filter(loiczid %in% macq_cells$loiczid) %>%
  inner_join(spp_ohi_iucn) %>%
  distinct() %>%
  group_by(iucn_sid, cat_score, sciname) %>%
  summarize(n_cells_spp = n()) %>%
  mutate(type = 'ohi')
i_rgn <- i_rgn_all %>%
  full_join(i_rgn_ohi)

a_rgn_overlap <- a_rgn %>% filter(is.na(type)) %>%
  ungroup() %>%
  mutate(weighted_cat = cat_score * n_cells_spp,
         weighted_cat_mean = sum(weighted_cat) / sum(n_cells_spp))
a_rgn_overlap$cat_score %>% mean()
a_rgn_overlap$n_cells_spp %>% mean()
a_rgn_overlap$weighted_cat_mean[1]
### Removed these species and replaced with IUCN:
### 0.1155556 mean cat score; 149.1556 mean cells; 45 species; mean weighted cat .2553

i_rgn_replace <- i_rgn %>% filter(sciname %in% a_rgn_overlap$sciname) %>%
  ungroup() %>%
  mutate(weighted_cat = cat_score * n_cells_spp,
         weighted_cat_mean = sum(weighted_cat) / sum(n_cells_spp))
i_rgn_replace$cat_score %>% mean()
i_rgn_replace$n_cells_spp %>% mean()
i_rgn_replace$weighted_cat_mean[1]
### These species overwrote AM species for combined score:
### 0.225 mean cat score; 260.2 mean cells; 16 spp; .191 mean weighted cat

look at weighted means globally

### 2016-08-17 ms_1_intro.Rmd

One major outcome of this body of science is the various compiled databases of species distribution maps. Two global-scale repositories predict marine species ranges throughout the world's oceans – AquaMaps [@kaschner_aquamaps:_2015] and International Union for Conservation of Nature (IUCN) [@iucn_iucn_2015].  These two spatial datasets have been used in hundreds of studies and applications for a wide range of purposes, including assessing marine species status [@halpern_index_2012; @halpern_patterns_2015; @selig_assessing_2013-1], evaluating global biodiversity patterns [@coll_biodiversity_2010; @martin_manual_2014; @pimm_biodiversity_2014; @kaschner_current_2011], predicting species range shifts [@garciamolinos_climate_2015], and setting conservation priorities [@klein_shortfalls_2015].

@@

<!--
The two datasets ostensibly describe the same information, but significant differences in methodology and intent could lead to dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations. 
Discuss "truth" and uncertainty in species distributions.
There is an objective "truth"; 
-->

Neither dataset claims to represent the "truth" of a species' spatial distribution, but rather each presents one understanding of the "truth" - one database relying primarily on expert opinion and the other on model predictions.  Uncertainties inherent in the method and intent inevitably drive differences in the range predictions made by each dataset; significant differences could indicate dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations.  Inaccurate indications of presence or absence could lead to ineffective marine reserve systems and management plans [@rondinini_tradeoffs_2006; @jetz_ecological_2008].  Importantly, biases in taxonomic or spatial coverage within a dataset could shift management and conservation actions away from places or species that are most in need.

To understand the implications of differences between the AquaMaps and IUCN datasets, we compare how each data source represents the global spatial and taxonomic distribution of marine species.  Most notably, AquaMaps includes range maps for many more species (currently `r format(n_spp_am, big.mark = ',')` species; `r p_spp_am`% of total), such that most global analyses related to marine biodiversity to date have used AquaMaps (IUCN range map data exist for only `r format(n_spp_iucn, big.mark = ',')` unique marine species).

___It seems fair to assume that/we can reasonably assume that___ range predictions from each dataset would be more closely aligned for well-understood species.  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align, determine several issues that lead to misalignment between predicted species distributions, and suggest methods to improve confidence in species range predictions.

>>>>

One major outcome of this body of science is the various compiled databases of species distribution maps. Here we focus on global-scale repositories that predict marine species ranges throughout the world's oceans – AquaMaps [@kaschner_aquamaps:_2015] and International Union for Conservation of Nature (IUCN) [@iucn_iucn_2015].  These two spatial datasets have been used in hundreds of studies and applications for a wide range of purposes, including assessing marine species status [@halpern_index_2012; @halpern_patterns_2015; @selig_assessing_2013-1], evaluating global biodiversity patterns [@coll_biodiversity_2010; @martin_manual_2014; @pimm_biodiversity_2014; @kaschner_current_2011], predicting species range shifts [@garciamolinos_climate_2015], and setting conservation priorities [@klein_shortfalls_2015].

@@

__Neither IUCN nor AquaMaps claims a lock on the "truth", and the small overlap in mapped species means each provides value in taxonomic and geographic coverage.  Method-driven differences in commission and omission errors produce clear and significant disagreement in species range descriptions between AquaMaps and IUCN datasets.__

Neither dataset claims to represent the "truth" of a species' spatial distribution, but rather each presents an understanding of the "truth" - one relying primarily on expert opinion and the other on model predictions, each method with its own challenges and advantages.  Uncertainties inherent in method and intent inevitably drive differences in the range predictions made by each dataset; significant differences could indicate dramatically different understandings of our marine ecosystems, with significant implications for policy and conservation recommendations.  Inaccurate indications of presence or absence could lead to ineffective marine reserve systems and management plans [@rondinini_tradeoffs_2006; @jetz_ecological_2008].  Importantly, biases in taxonomic or spatial coverage within a dataset could shift management and conservation actions away from places or species that are most in need.

To understand the implications of differences between the AquaMaps and IUCN datasets, we compare how each data source represents the global spatial and taxonomic distribution of the `r format(n_spp_all, big.mark = ',')` marine species mapped by one or both datasets.  Most notably, AquaMaps includes range maps for many more species (currently `r format(n_spp_am, big.mark = ',')` species; `r p_spp_am`% of total), such that most global analyses related to marine biodiversity to date have used AquaMaps (IUCN range map data exist for only `r format(n_spp_iucn, big.mark = ',')` unique marine species).

As understanding of a given species improves, we expect that range maps, regardless of method, would become more accurate in predicting both species presence and absence.  Close alignment between two independent range maps can therefore indicate higher confidence in species presence or absence than a single map.  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align in both distribution and overall area; from the results we determine several issues that can introduce commission errors, and suggest methods to improve confidence in species range predictions.
